---
title: "An introduction functions in R"
subtitle: "https://bradduthie.github.com/talks/intro_to_R.pdf"
author: "Brad Duthie (alexander.duthie@stir.ac.uk)"
date: "16 November 2022"
output:
  beamer_presentation: default
header-includes:
- \usepackage{hyperref}
colorlinks: true
linkcolor: blue
urlcolor: blue
---


# What is a function?

\Large

**Functions are ubiquitous and highly useful in R**

\begin{itemize}
\setlength\itemsep{0.4em}
\item Code that is organised to perform a specific task$^{1}$
\item Can get functions in multiple ways
\begin{enumerate} \Large
\item R base functions
\item R package functions
\item Custom functions
\end{enumerate}
\item All functions have a similar structure
\end{itemize}

\footnotetext[1]{R-Functions. Tutorialspoint. https://www.tutorialspoint.com/r/r\_functions.htm Accessed 13 NOV 2022.}


# What is a function?


```{r, eval = FALSE}
function_name <- function(arg1, arg2 = default){
  # Code that can use arguments, arg1 & arg2
  return(output);
}
```

\vspace{2mm}
\hrule \pause
\vspace{2mm}

```{r}
my_mean <- mean(x = 1:10);
```

```{r, echo = FALSE}
print(my_mean);
```

\vspace{2mm}
\hrule \pause
\vspace{2mm}

```{r}
class(mean);
```


# Base functions in R

\LARGE

**Base R includes hundreds of functions**

\begin{itemize}
\setlength\itemsep{0.0em}
\item Most \href{https://stat.ethz.ch/R-manual/R-devel/library/base/html/00Index.html}{base functions} not used$^{1}$
\item Familiar functions \texttt{mean}, \texttt{plot}, \texttt{summary}
\item Includes functions like \texttt{+}, \texttt{<-}, \texttt{"}, or \texttt{!}
\end{itemize}

Additional functions can be found in R packages, or custom made and read into the R console.

\footnotetext[1]{https://stat.ethz.ch/R-manual/R-devel/library/base/html/00Index.html}





# Non-base functions in R

**Functions outwith base R available in packages**

\begin{itemize}
\setlength\itemsep{0.0em}
\item \href{https://cran.r-project.org/}{Comprehensive R Archive Network} includes 18000+ packages
\item Packages include specialised functions
\item Access with `install.packages` and `library`
\end{itemize}


```{r, eval = FALSE}
install.packages("ggplot2");
library("ggplot2");
ggplot(data = dat, mapping = aes(x = wgt, y = totlen))
       + geom_point();
```


Custom functions can be written in R too with the `function` function.


# A custom function in R

Convert from Fahrenheit to Celsius


```{r}
F_to_C <- function(F_temp){
    C_temp <- (F_temp - 32) * 5/9;
    return(C_temp);
}
```

Highlight the whole function and run it, then you can use it.


```{r}
F_to_C(F_temp = 70);
```

Now write a custom function for C to F!



# Functions within functions

**We can use a custom function within another custom function.**

Convert from Fahrenheit to Kelvin.

```{r}
F_to_K <- function(F_temp){
  K_temp <- F_to_C(F_temp = F_temp) + 273.15;
  return(K_temp);
}
```

Because Kelvin equals degrees Celsius plus 273.15, we can call `F_to_C`, then add 273.15 to it.









